\subsection{Agda Mechanization}
\label{sec:agda}

Our described system is fairly intricate, with many of the proofs
requiring significant case-work.
%
For instance, while \autoref{theorem:exhaustive-truify} and
\autoref{theorem:redundant-truify-falsify} are intuitively correct, it
took us several attempts before ending up with separating the match
constraint language presented in the paper into a two-level system.
%
\todo{hope this highlights the difficulty/contribution}
%
Thus, to ensure
no details have been overlooked, the supplementary material contains a
mechanization of the semantics and metatheory of Peanut using the Agda
proof assistant \cite{norell:thesis}. This includes all of the
theorems and lemmas stated above and in the appendix.

In general, our mechanization takes an approach similar to that used in the mechanization of Hazelnut Live \cite{DBLP:journals/pacmpl/OmarVCH19}. As is typical in Agda, judgments are encoded as inductive datatypes, and their rules are encoded as dependently typed constructors thereof. For variable and hole names, note that on paper, we ignore issues related to shadowing, and implicitly assume that we may $\alpha$-convert terms as needed. In our mechanization, rather than explicitly performing such renaming, we insert appropriate uniqueness constraints as a premise to each theorem. Additionally, we make assumptions about disjointedness between bound variables and typing contexts, following a slightly generalized version of Barendregt’s convention \cite{DBLP:books/daglib/0067558, DBLP:conf/cade/UrbanBN07}. All such premises are fairly benign, as they can always be satisfied by some $\alpha$-equivalent term.

Furthermore, again following the mechanization of Hazelnut Live, we restrict variable names to natural numbers, and we encode typing contexts and hole contexts as metafunctions $\mathbb{N} \to \text{Maybe } T$ for appropriate types $T$. While not strictly necessary, we postulate function extensionality to improve the ergonomics of such an encoding. This postulate is fairly innocuous, as function extensionality is known to be independent of Agda’s axioms \cite{DBLP:conf/lics/AwodeyGS12}. We assume no other postulates.

In a few places, the mechanization also differs slightly from the definitions laid out on paper. Most notably, the mechanization removes exhaustiveness and redundancy checking from the typing judgment, and instead includes them separately as their own judgments. This has the advantage of making these checks optional, while also avoiding a morally-irrelevant but technically-inconvenient positivity issue with the encoding of the typing judgment in Agda. As a consequence, we also prove separate preservation theorems for each of these judgments, and take exhaustiveness as an additional assumption in the Progress theorem. The documentation provided with the mechanization has more details, enumerating all deviations from the paper and their motivation.