\section{Labeled Sums}\label{sec:labeledsums}

In this section, we conservatively extend Peanut with finite labeled sums, which is a practical necessity in general-purpose functional languages.

The first step is to generalize the syntax of binary sums to more than two variants:
\[
  \arraycolsep=4pt\begin{array}{lll}
    \tau & ::= & \ldots ~\vert~ \tlabeledsum{C_i(\tau_i)}_{C_i \in \tagset} \\
    e    & ::= & \ldots ~\vert~ \hinj{C}{\tau}{e}                           \\
    p    & ::= & \ldots ~\vert~ \hinjp{C}{p}                                \\
    C    & ::= & \tagC ~\vert~ \tagehole{u} ~\vert~ \taghole{\tagC}{u}
  \end{array}
\]
Labeled sums introduce a new sort $C$ for labels, a.k.a. datatype constructors, and a new type-level connective $C_1(\tau_1) + \cdots + C_n(\tau_n)$ for gathering and labeling types.
Since we are not usually concerned with the length of any particular sum,
we adopt a slightly more general and compact notation $\tlabeledsum{C_i(\tau_i)}_{C_i \in \tagset}$
for the sum consisting of labels $\tagset = \setof{C_i}_{i \leq n}$ with respective argument types $\setof{\tau_i}_{i \leq n}$.
The introduction form is the \emph{labeled injection expression} $\hinj{C}{\tau}{e}$ for injecting expression $e$ into sum $\tau$ at label $C$.
The elimination form is the \emph{labeled injection pattern} $\hinj{C}{\tau}{p}$ 
for expressions matching pattern $p$ that have been injected into sum $\tau$ at label $C$.
% For the remainder of this section, we assume all injections are labeled unless otherwise stated,
% and we elide sorts that are irrelevant or implied by the context.

To ensure maximal liveness, we distinguish \emph{concrete labels} $\tagC$
from \emph{label holes} which are either empty $\tagehole{u}$ or not empty $\taghole{\tagC}{u}$.
Empty holes arise where labels have yet to be constructed, for example during incremental construction of a sum or injection.
Non-empty holes operate as membranes around labels that are syntactically malformed or that violate a semantic constraint.
Non-empty holes around syntactically valid labels indicate duplication for sum type declarations and non-membership for injections.

\input{fig-labeled-sums-pattern-matching-rules}
Moving on to the dynamic semantics, \Autoref{fig:labeled-sums-pattern-matching-rules,fig:labeled-sums-evaluation-rules}
extend Peanut's dynamic semantics to support evaluation of labeled injections.
The rules in \autoref{fig:labeled-sums-pattern-matching-rules} define pattern matching on injections with label holes,
subject to the matching determinism conditions imposed by \autoref{lemma:match-determinism}.
Rule \MInj is a straightforward rule for matching labeled injections.
Rules \MMInjTag and \MMInjArg allow indeterminate matching against a pattern whose label or argument indeterminately match, respectively.
Rule \NMInj forbids matching of concrete but unequal labels, and
Rules \NMInjTag and \NMInjArg forbid matching when the arguments do not match.
Rules \TMMSym, \TMMHole, and \TMMEHole define indeterminate matching of labels
by establishing a partial equivalence among distinct labels when at least one of them is a hole.
\autoref{fig:labeled-sums-evaluation-rules} defines stepping of injection expressions,
subject to the determinism conditions imposed by \autoref{theorem:determinism}.
These rules are direct adaptations of the corresponding rules for binary injections in \Autoref{fig:final,fig:step}.
\input{fig-labeled-sums-evaluation-rules}

\input{fig-labeled-sums-typing-rules}
Moving on to the static semantics, we first extend the typing relations as shown in \autoref{fig:labeled-sums-typing-rules}
by generalizing the rules for binary injections from \Autoref{fig:exptyp,fig:pat-rulestyp}.
The rules shown in \autoref{fig:labeled-sums-typing-rules} are essentially the same as their binary counterparts,
differing in that instead of assuming all sums contain precisely two fixed labels,
we must check whether label $C_j$ with argument type $\tau_j$ belongs to the annotated sum type $\tau$.

\input{fig-labeled-sums-constraint-rules}
Next, we extend the syntax of constraints to labeled sums:
\[
  \arraycolsep=4pt\begin{array}{lll}
    \xi & ::= & \ldots ~\vert~ \cinj{C}{\tau}{\xi}
  \end{array}
\]
\autoref{fig:labeled-sums-constraint-rules} extends Peanut's constraint checking semantics to support matching on injections into labeled sums.
A constraint of the form $\cinj{C}{\tau}{\xi}$ represents injections into sum $\tau$ with label $C$ with arguments constrained by $\xi$.
Rule \CTInj decides which sum type is constrained by an injection constraint.
Rules \CSInj, \CMSInjTag, and \CMSInjArg specify whether and to what extent injections satisfy injection constraints,
subject to the coherence conditions imposed by \autoref{theorem:exclusive-constraint-satisfaction} and \autoref{lemma:const-matching-coherence}.
That is, an injection satisfies a constraint only when their labels are equal and the argument expression satisfies the argument constraint.

\input{fig-labeled-sums-refutability-rules}
\input{fig-labeled-sums-decidability}
Finally, we extend the inconsistency relations as shown in \Autoref{fig:labeled-sums-refutability-rules,fig:labeled-sums-decidability}.
Rule \RInjMult establishes that constraints for nontrivial sums are refutable, i.e., that the set of dual constraints is not empty.
The dual of a constraint with a concrete label consists of injections of the same label with dual arguments as well as injections of other labels into the sum.
A constraint with a label hole is dual to all injections into the sum with concrete labels.
The premises of Rule \RInjSing merely express the fact that there are no alternatives injections into singleton sums, and none at all into void sums.
Rule \PTInj establishes that it is possible to satisfy a labeled injection constraint whenever it is possible to satisfy its argument constraint.
Rule \CINCInjTag detects inconsistent labels.
Since label holes may match any other label, it suffices to consider only concrete labels.
Rule \CINCInjArg detects inconsistent argument constraints, regardless of label status.
Both rules are necessary to ensure the absence of inconsistent injection constraints.

% The inclusion of finite labeled sums with label holes adds substantial complexity to type consistency checking.
% For example, given two finite labeled sum types of cardinality $n$,
% whether they are consistent depends on whether there is a consistent pairing of the elements.
% Naively generating all $n^2$ pairings and then checking every subset of size $n$ 
% would require $\binom{n^2}{n} \gg n^2$ checks.
% If we instead model the search space as a complete bipartite graph with the elements of the sums on opposing sides,
% the problem reduces to one of finding a maximum matching in a bipartite graph with $2n$ vertices and $n^2$ edges,
% which can be solved by Hopcroft-Karp\todo{cite} in $O(n^{5/2})$ steps.
